A planetary collision afterglow and transit of the resultant debris cloud

### 2023-10-12 src/tex/ms.tex
[showyourwork] overleaf sync

\affil[19]{\orgdiv{Department of Astronomy}, \orgname{The Ohio State University}, \orgaddress{\street{140 West 18th Avenue}, \city{Columbus}, \postcode{43210}, \state{OH}, \country{USA}}}

\abstract{
Planets grow in rotating disks of dust and gas around forming stars, some of which can subsequently collide in giant impacts after the gas component is removed from the disk \cite{Williams11,Wyatt15,Hughes18}.
%
Monitoring programs with the warm Spitzer mission have recorded significant and rapid changes in mid-infrared output for several stars, interpreted as variations in the surface area of warm dusty material ejected by planetary-scale collisions and heated by the central star: e.g., NGC 2354–ID8 \cite{2014Sci...345.1032M,Su19}, HD 166191 \cite{Su22} and V844 Persei \cite{2021ApJ...918...71R}.
%
Here we report combined observations of the young ($\sim$~300~Myr), solar-like star \asas{}: an infrared brightening consistent with a blackbody temperature of 1000 K and a luminosity of 4\% of that of the star lasting for about 1000\,days, partially overlapping in time with a complex and deep wavelength-dependent optical eclipse that lasted for about 500 days.
%
The optical eclipse started 2.5\,years after the infrared brightening, implying an orbital period of at least that duration.
%
These observations are consistent with a collision between two exoplanets of several to tens of Earth masses at 2-16\,au from the central star. 
%
Such an impact produces a hot, highly-extended post-impact remnant with sufficient luminosity to explain the infrared observations.
%
Transit of the impact debris, sheared by orbital motion into a long cloud, causes the subsequent complex eclipse of the host star.
}

@@

% ▀█▀ █░█ █▀▀   █▀▀ █░█ █▀▀ █▄░█ ▀█▀
% ░█░ █▀█ ██▄   ██▄ ▀▄▀ ██▄ █░▀█ ░█░

@@

Giant impacts produce significant amounts of debris, typically around 1\% of the colliding mass, that is injected into orbit around the host star\cite{Canup2001,Lock18}.
%
Impact ejecta has a wide range of sizes, from sub-micron dust to planetesimals of tens to hundreds of kilometers, and often contains the most highly heated material \cite{Benz2008_Mercury_book,Leinhardt2015,Carter2020a}.
%
For sufficiently high impact velocities ($>1$~km~s$^{-1}$ for water ice, and $>8$~km~s$^{-1}$ for forsterite) a substantial fraction of this material is vaporised \cite{Stewart2008,Davies2020,Carter2020a}.
%
Shearing of droplets and cooling and condensation of vapor produces a population of small dust grains and solid spherules.
%
The size distribution of this fraction of the debris is uncertain, due to difficulties in modelling condensate nucleation and breakup, but previous work suggests that debris could range in size from sub-micron to decimeters \cite{Benz2008_Mercury_book,Johnson2015}.
%
The wavelength-dependent eclipse suggests that the optical depth of the transiting dust cloud is dominated by sub-micron grains, consistent with these previous estimates.
%
While the fading of the excess infrared flux occurs within 100 days of the start of the optical transit, we consider the timing to be coincidental because there is no clear correspondence between the light curves, for example no  change in the (six month cadence) IR flux when the star dims in the optical wavelengths just before MJD 59500.

@@

Independent of impact simulations, there are fundamental limits on the size of post-impact bodies from the Hill and Bondi radii.
%
Any post-impact structure must lie within the Hill sphere, the distance within which the gravity of an object dominates over that of the star.
%
Furthermore, only vapor within the Bondi radius would be bound to the post-impact body.
%
Figure~\ref{fig:Hill_Bondi_R}A shows the Hill radii at different distances from ASASSN-21qj (solid lines) and the Bondi radius for example gas species (dotted lines).
%
Beyond 2.4\,au, the Hill radii of greater than Earth-mass bodies are large enough to accommodate a post-impact body capable of producing the required IR flux ($\sim7R_*$, dotted line). 
%
Heavier gases (H$_2$O, SiO, and SiO$_2$) are also bound to bodies of more than a few Earth masses.
%
A post-impact body of a few Earth masses can hence theoretically produce the observed IR emission, with a photosphere that is dust and/or vapor.

@@

The temporal variation of flux from a post-impact body is controlled by evolution of the bodies' size which is governed by a number of competing factors \cite{Lock2018moon,Lock2020}, including radiative energy loss, viscous spreading, and mass and angular momentum transfer by condensates.
%
For silicate-dominated post-impact bodies, the high emission temperature means that radiative cooling dominates, and the post-impact body contracts rapidly, fully condensing over years to thousands of years \cite{Lock2018moon,Lock2020}.
%
The body we observed has a much lower emission temperature, and contracted substantially more slowly.
%
Figure~\ref{fig:Hill_Bondi_R}~B and C show evolution of conceptual post-impact bodies in the limiting case that radiative cooling dominates.
%
If sufficient mass is injected into the outer regions (i.e., less negative power law exponents) the observed flux can remain constant for an initial period and then decays over the order of hundreds of days, in agreement with infrared observations.
%
Further work is required to understand the structure and evolution of bodies produced by impacts between super-Earths and mini-Neptunes of different compositions and identify temporal flux variations consistent with our observations.

@@

The stellar properties of \asas\ (Gaia DR3 5539970601632026752 = 2MASS~J08152329-3859234) are listed in Extended Data Table~\ref{tab:Stellarprop}, showing that \asas\ is consistent with being a G2 type dwarf star.
%
Where necessary we assume a stellar mass equal to the Sun.
%
\asas\  has a neighbor (Gaia DR3 5539970597334497024 = 2MASS~J08152298-3859244) which is a visual double.
%
Based on the Gaia EDR3 mean ICRS position for epoch 2016.0, the visual companion lies at a separation $\rho$ = $3738.243\pm0.062$ mas and at position angle $\theta$ = $249^{\circ}.977$.
%
Their parallaxes ($\varpi$ = $1.7631\pm0.0112$ mas vs. $1.4711\pm0.0523$ mas) differ by 5.5$\sigma$ and proper motions ($\mu_{\alpha} = -9.692\pm0.012$, $\mu_{\delta} = 7.349\pm0.012$ \masyr\, vs. $\mu_{\alpha} = -0.114\pm0.055$, $\mu_{\delta} = 6.419\pm0.053$ \masyr) differ by a factor of 2.
%
The large differences in distance and proper motions suggests that these stars are not associated.

@@

\begin{table}
    \centering
    \caption{Photometric observations of ASASSN-21qj. The number of points are listed per survey and filter. %
    %
    This is the count after initial rejection of photometric points with significantly large error bars.
    %
    Further photometric points may have been rejected in the differeny analysis steps.
    %
    The ROAD observations constitute the majority of observations from the AAVSO data sets.}
    \begin{tabular}{@{}lcc@{}}
    \hline\hline
Survey name  & Filter & Number of points                    \\
        \hline
ASASSN   &  $V$   & 758 \\
   &  $g'$   & 3225 \\
        \hline
ATLAS & $c$ & 161 \\
 & $o$ & 677 \\
        \hline
AAVSO   &  $B$   & 729 \\    % 601 in RVC
   &  $I$   & 728 \\    % 612 in RVC
   &  $V$   & 722 \\    % 599 in RVC
        \hline
LCOGT   &  $g$   & 275 \\    % RVC
   &  $r$   & 224 \\    % RVC
   &  $i$   & 168 \\    % RVC
        \hline
NEOWISE & $W1$ & 18 \\
 & $W2$ & 18 \\
        \hline
    \end{tabular}

@@

How the emission from a post-impact body would evolve with time is highly dependent on the initial mass distribution and thermal state of the body, and the balance between radiative cooling, viscous spreading, and mass and angular momentum transport by condensates \cite{Lock2017,Lock2018moon,Lock2020}. 
%
Given the limitations of SPH simulations (see above) it is not possible to accurately determine the initial structure of post-impact bodies in the relevant regime.
%
To explore a range of possible evolution pathways, we have calculated the evolution of post-impact bodies with different power-law surface density profiles, $\Sigma$, profiles, under the limiting case that radiative cooling and condensation of the vapor is the single driver for evolution of the structure. 
%
We chose a power-law surface density profile as it can straightforwardly cover the wide range of surface density profiles expected after super-Earth/mini-Neptune collisions, based on that found in impact simulations between lower-mass, terrestrial bodies \cite{Lock2017,Canup2001,Canup2012,Cuk2012,Rufu2017,Reufer2012}. 
%
The surface density profile is given by:
%
\begin{equation}
\Sigma=\begin{cases}
\Sigma_0, & \text{if } r_{xy}\leq R_{\rm c}\\
\alpha r_{xy}^{\beta}, & \text{if } R_{\rm c} > r_{xy}\leq R_{\rm emit}\\
0, & \text{if } r_{xy}> R_{\rm emit}
\end{cases}   
\label{eqn:sigma}
\end{equation}
%
where: $r_{xy}$ is the distance from the rotation axis; $R_{\rm c}$ is the outer radius of a constant surface density central region, roughly analogous to the corotating regions seen in Earth-mass synestias \cite{Lock2017,Lock2018moon}; $\Sigma_0$ is the surface density of the central region; $\beta$ is the power law exponent; and $R_{\rm emit}$ is the initial emitting radius. Imposing surface density continuity gives
%
\begin{equation}
    \alpha=\Sigma_0 R_{\rm c}^{-\beta} \; .
\end{equation}
%
We can determine $\Sigma_0$ by fixing the mass of the body, 
%
\begin{equation}
    M_{\rm p} = \int_0^{R_{\rm c}}{2 \pi r'_{xy} \Sigma_0 dr'_{xy}} + \int_{R_{\rm c}}^{R_{\rm emit}}{2 \pi r'_{xy} \Sigma(r'_{xy}) dr'_{xy}} \; ,
\end{equation}
%
which can be solved to give:
%
\begin{equation}
\Sigma_0=\begin{cases}
\frac{M_{\rm p}}{\pi \left [R_{\rm c}^2 + 2 R_{\rm c}^{-\beta} \left ( \ln{R_{\rm emit}}-\ln{R_{\rm c}}\right ) \right ]}, & \text{if } \beta=-2,\\
\frac{M_{\rm p}}{\pi \left [R_{\rm c}^2 + \frac{2 R_{\rm c}^{-\beta}}{\beta+2} \left ( R_{\rm emit}^{\beta+2}-R_{\rm c}^{\beta+2}\right ) \right ]}, & \text{otherwise.} 
\end{cases}
\end{equation}
%
The time taken for a given region of the structure to cool to the point that a sufficient fraction of material is condensed for the temperature to drop below the condensation-buffer (see above) and so the emitted flux to drop is given by
%
\begin{equation}
    t_{\rm cool} (r_{xy})=\frac{l f(r_{xy}) \Sigma(r_{xy})}{\sigma T^4_{\rm emit}}
\end{equation}
%
where $f$ is the initial vapor fraction at that radius, $l$ is the latent heat of vaporization of the material (here we have taken the limiting case of pure water $l=2.256\times 10^6$ \cite{Chase1998}, but the addition of silicates could make the latent heat much larger), $\sigma$ is the Stefan–Boltzmann constant, and $T_{\rm emit}$ is the emission temperature.
%
Figure~\ref{fig:Hill_Bondi_R} shows example evolutions of the emission from a post-impact body using this model for different parameters (solid lines).
%
Results of a modified model where the initial surface density (by the addition of a constant parameter to Equation~\ref{eqn:sigma}) is forced to be zero at $R_{\rm emit}$ are shown as dashed lines.

>>>>

\affil[19]{\orgdiv{Department of Astronomy}, \orgname{The Ohio State University}, \orgaddress{\street{140 West 18th Avenue}, \city{Columbus}, \postcode{43210}, \state{OH}, \country{USA}}}
%
% Adding empty lines to make line count the same as ms-edited.tex
%
\abstract{
Planets grow in rotating disks of dust and gas around forming stars, some of which can subsequently collide in giant impacts after the gas component is removed from the disk \cite{Williams11,Wyatt15,Hughes18}.

Monitoring programs with the warm Spitzer mission have recorded significant and rapid changes in mid-infrared output for several stars, interpreted as variations in the surface area of warm dusty material ejected by planetary-scale collisions and heated by the central star: e.g., NGC 2354–ID8 \cite{2014Sci...345.1032M,Su19}, HD 166191 \cite{Su22} and V844 Persei \cite{2021ApJ...918...71R}.
%
Here we report combined observations of the young ($\sim$~300~Myr), solar-like star \asas{}: an infrared brightening consistent with a blackbody temperature of 1000 K and a luminosity of 4\% of that of the star lasting for about 1000\,days, partially overlapping in time with a complex and deep wavelength-dependent optical eclipse that lasted for about 500 days.
%
The optical eclipse started 2.5\,years after the infrared brightening, implying an orbital period of at least that duration.
%
These observations are consistent with a collision between two exoplanets of several to tens of Earth masses at 2-16\,au from the central star. 
%
Such an impact produces a hot, highly-extended post-impact remnant with sufficient luminosity to explain the infrared observations.
%
Transit of the impact debris, sheared by orbital motion into a long cloud, causes the subsequent complex eclipse of the host star.
}

@@

% ▀█▀ █░█ █▀▀   █▀▀ █░█ █▀▀ █▄░█ ▀█▀
% ░█░ █▀█ ██▄   ██▄ ▀▄▀ ██▄ █░▀█ ░█░
%
%
%
%
%
%
% Adding empty lines to make line count the same as ms-edited.tex
% Makes it easier to crosscheck
%
%
%
%

@@

Giant impacts produce significant amounts of debris, typically around 1\% of the colliding mass, that is injected into orbit around the host star\cite{Canup2001,Lock18}.
%
Impact ejecta have a wide range of sizes, from sub-micron dust to planetesimals of tens to hundreds of kilometers, and often contains the most highly heated material \cite{Benz2008_Mercury_book,Leinhardt2015,Carter2020a}.
%
For sufficiently high impact velocities ($>1$~km~s$^{-1}$ for water ice, and $>8$~km~s$^{-1}$ for forsterite) a substantial fraction of this material is vaporised \cite{Stewart2008,Davies2020,Carter2020a}.
%
Shearing of droplets and cooling and condensation of vapor produces a population of small dust grains and solid spherules.
%
The size distribution of this fraction of the debris is uncertain, due to difficulties in modelling condensate nucleation and breakup, but previous work suggests that debris could range in size from sub-micron to decimeters \cite{Benz2008_Mercury_book,Johnson2015}.
%
The wavelength-dependent eclipse suggests that the optical depth of the transiting dust cloud is dominated by sub-micron grains, consistent with these previous estimates.
%
While the fading of the excess infrared flux occurs within 100 days of the start of the optical transit, we consider the timing to be coincidental because there is no clear correspondence between the light curves, for example no  change in the (six month cadence) IR flux when the star dims in the optical wavelengths just before MJD 59500.

@@

Independent of impact simulations, there are fundamental limits on the size of post-impact bodies from the Hill and Bondi radii.
%
Any post-impact structure must lie within the Hill sphere, the distance within which the gravity of an object dominates over that of the star.
%
Furthermore, only vapor within the Bondi radius would be bound to the post-impact body.
%
Figure~\ref{fig:Hill_Bondi_R}A shows the Hill radii at different distances from ASASSN-21qj (solid lines) and the Bondi radii for example gas species (dotted lines).
%
Beyond 2.4\,au, the Hill radii of greater than Earth-mass bodies are large enough to accommodate a post-impact body capable of producing the required IR flux ($\sim7R_*$, dotted line). 
%
Heavier gases (H$_2$O, SiO, and SiO$_2$) are also bound to bodies of more than a few Earth masses.
%
A post-impact body of a few Earth masses can hence theoretically produce the observed IR emission, with a photosphere that is dust and/or vapor.

@@

The temporal variation of flux from a post-impact body is controlled by evolution of the bodies' size which is governed by a number of competing factors \cite{Lock18,Lock2020}, including radiative energy loss, viscous spreading, and mass and angular momentum transfer by condensates.
%
For silicate-dominated post-impact bodies, the high emission temperature means that radiative cooling dominates, and the post-impact body contracts rapidly, fully condensing over years to thousands of years \cite{Lock18,Lock2020}.
%
The body we observed has a much lower emission temperature, and contracted substantially more slowly.
%
Figure~\ref{fig:Hill_Bondi_R}~B and C show evolution of conceptual post-impact bodies in the limiting case that radiative cooling dominates.
%
If sufficient mass is injected into the outer regions (i.e., less negative power law exponents) the observed flux can remain constant for an initial period and then decays over the order of hundreds of days, in agreement with infrared observations.
%
Further work is required to understand the structure and evolution of bodies produced by impacts between super-Earths and mini-Neptunes of different compositions and identify temporal flux variations consistent with our observations.

@@

The stellar properties of \asas\ (Gaia DR3 5539970601632026752 = 2MASS~J08152329-3859234) are listed in Extended Data Table~\ref{tab:Stellarprop}, showing that \asas\ is consistent with being a G2 type dwarf star.
%
Where necessary we assume a stellar mass equal to the Sun.
%
\asas\  has a neighbor (Gaia DR3 5539970597334497024 = 2MASS~J08152298-3859244) which is a visual double.
%
Based on the Gaia DR3 mean ICRS position for epoch 2016.0, the visual companion lies at a separation $\rho$ = $3738.243\pm0.062$ mas and at position angle $\theta$ = $249^{\circ}.977$.
%
Their parallaxes ($\varpi$ = $1.7631\pm0.0112$ mas vs. $1.4711\pm0.0523$ mas) differ by 5.5$\sigma$ and proper motions ($\mu_{\alpha} = -9.692\pm0.012$, $\mu_{\delta} = 7.349\pm0.012$ \masyr\, vs. $\mu_{\alpha} = -0.114\pm0.055$, $\mu_{\delta} = 6.419\pm0.053$ \masyr) differ by a factor of 2.
%
The large differences in distance and proper motions suggests that these stars are not associated.

@@

\begin{table}
    \centering
    \caption{Photometric observations of ASASSN-21qj. The number of points are listed per survey and filter. %
    %
    This is the count after initial rejection of photometric points with significantly large error bars.
    %
    Further photometric points may have been rejected in the different analysis steps.
    %
    The ROAD observations constitute the majority of observations from the AAVSO data sets.}
    \begin{tabular}{@{}lcc@{}}
    \hline\hline
Survey name  & Filter & Number of points                    \\
        \hline
ASASSN   &  $V$   & 758 \\
   &  $g'$   & 3225 \\
        \hline
ATLAS & $c$ & 161 \\
 & $o$ & 677 \\
        \hline
AAVSO   &  $B$   & 729 \\    % 601 in RVC
   &  $I$   & 728 \\    % 612 in RVC
   &  $V$   & 722 \\    % 599 in RVC
        \hline
LCOGT   &  $g$   & 275 \\    % RVC
   &  $r$   & 224 \\    % RVC
   &  $i$   & 168 \\    % RVC
        \hline
NEOWISE & $W1$ & 18 \\
 & $W2$ & 18 \\
        \hline
    \end{tabular}

@@

How the emission from a post-impact body would evolve with time is highly dependent on the initial mass distribution and thermal state of the body, and the balance between radiative cooling, viscous spreading, and mass and angular momentum transport by condensates \cite{Lock2017,Lock18,Lock2020}. 
%
Given the limitations of SPH simulations (see above) it is not possible to accurately determine the initial structure of post-impact bodies in the relevant regime.
%
To explore a range of possible evolution pathways, we have calculated the evolution of post-impact bodies with different power-law surface density profiles, $\Sigma$, profiles, under the limiting case that radiative cooling and condensation of the vapor is the single driver for evolution of the structure. 
%
We chose a power-law surface density profile as it can straightforwardly cover the wide range of surface density profiles expected after super-Earth/mini-Neptune collisions, based on that found in impact simulations between lower-mass, terrestrial bodies \cite{Lock2017,Canup2001,Canup2012,Cuk2012,Rufu2017,Reufer2012}. 
%
The surface density profile is given by:
%
\begin{equation}
\Sigma=\begin{cases}
\Sigma_0, & \text{if } r_{xy}\leq R_{\rm c}\\
\alpha r_{xy}^{\beta}, & \text{if } R_{\rm c} > r_{xy}\leq R_{\rm emit}\\
0, & \text{if } r_{xy}> R_{\rm emit}
\end{cases}   
\label{eqn:sigma}
\end{equation}
%
where: $r_{xy}$ is the distance from the rotation axis; $R_{\rm c}$ is the outer radius of a constant surface density central region, roughly analogous to the corotating regions seen in Earth-mass synestias \cite{Lock2017,Lock18}; $\Sigma_0$ is the surface density of the central region; $\beta$ is the power law exponent; and $R_{\rm emit}$ is the initial emitting radius. Imposing surface density continuity gives
%
\begin{equation}
    \alpha=\Sigma_0 R_{\rm c}^{-\beta} \; .
\end{equation}
%
We can determine $\Sigma_0$ by fixing the mass of the body, 
%
\begin{equation}
    M_{\rm p} = \int_0^{R_{\rm c}}{2 \pi r'_{xy} \Sigma_0 dr'_{xy}} + \int_{R_{\rm c}}^{R_{\rm emit}}{2 \pi r'_{xy} \Sigma(r'_{xy}) dr'_{xy}} \; ,
\end{equation}
%
which can be solved to give:
%
\begin{equation}
\Sigma_0=\begin{cases}
\frac{M_{\rm p}}{\pi \left [R_{\rm c}^2 + 2 R_{\rm c}^{-\beta} \left ( \ln{R_{\rm emit}}-\ln{R_{\rm c}}\right ) \right ]}, & \text{if } \beta=-2,\\
\frac{M_{\rm p}}{\pi \left [R_{\rm c}^2 + \frac{2 R_{\rm c}^{-\beta}}{\beta+2} \left ( R_{\rm emit}^{\beta+2}-R_{\rm c}^{\beta+2}\right ) \right ]}, & \text{otherwise.} 
\end{cases}
\end{equation}
%
The time taken for a given region of the structure to cool to the point that a sufficient fraction of material is condensed for the temperature to drop below the condensation-buffer (see above) and so the emitted flux to drop is given by
%
\begin{equation}
    t_{\rm cool} (r_{xy})=\frac{l f(r_{xy}) \Sigma(r_{xy})}{\sigma T^4_{\rm emit}}
\end{equation}
%
where $f$ is the initial vapor fraction at that radius, $l$ is the latent heat of vaporization of the material (here we have taken the limiting case of pure water $l=2.256\times 10^6$ \cite{Chase1998}, but the addition of silicates could make the latent heat much larger), $\sigma$ is the Stefan–Boltzmann constant, and $T_{\rm emit}$ is the emission temperature.
%
Figure~\ref{fig:Hill_Bondi_R} shows example evolutions of the emission from a post-impact body using this model for different parameters (solid lines).
%
Results of a modified model where the initial surface density (by the addition of a constant parameter to Equation~\ref{eqn:sigma}) is forced to be zero at $R_{\rm emit}$ are shown as dashed lines.
